---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/images/README-",
  out.width = "100%"
)

# save the built-in output hook
hook_output <- knitr::knit_hooks$get("output")
hook_message <- knitr::knit_hooks$get("message")

# define a truncation helper
truncate_lines_tail <- function(x, lines) {
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines) == 1) {
    if (length(x) > lines) {
      # truncate the output, but add `more`
      # x <- c(head(x, n = lines), more)
      x <- c(more, tail(x, n = lines))
    }
  } else {
    x <- c(more, x[lines], more)
  }
  paste(c(x, ""), collapse = "\n")
}

# set the new hooks
knitr::knit_hooks$set(
  output = function(x, options) {
    lines <- options$out.lines
    if (is.null(lines)) return(hook_output(x, options))
    x <- truncate_lines_tail(x, lines)
    return(hook_output(x, options))
  },
  message = function(x, options) {
    lines <- options$out.lines
    if (is.null(lines)) return(hook_message(x, options))
    x <- truncate_lines_tail(x, lines)
    return(hook_message(x, options))
  }
)

# helper function to get the number of rows of an RO-Crate
rocrate_lines <- function(rocrate) {
  rocrate_txt(rocrate) |>
    length()
}

# helper function to read RO-Crate as text file
rocrate_txt <- function(rocrate) {
  # create tmp file to write a JSON file with the output
  tmp_file <- tempfile(fileext = ".json")
  on.exit(unlink(tmp_file, force = TRUE))
  # write RO-Crate in tmp file
  rocrateR::write_rocrate(rocrate, tmp_file)
  # read RO-Crate as text file
  readLines(tmp_file)
}
```

# dsROCrate: 'DataSHIELD' RO-Crate Wrapper Functions

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/dsROCrate)](https://CRAN.R-project.org/package=dsROCrate)
<!-- badges: end -->

The goal of dsROCrate is to provide functions to wrap elements from a 
'DataSHIELD' analysis into an RO-Crate.

## 1. Installation

You can install the development version of dsROCrate from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("DataSHIELD-5S/dsROCrate")
```

## 2. Creating your first RO-Crate
In this example, we will be using OBiBa's 
[Opal](https://opaldoc.obiba.org/en/latest/index.html) as the _back-end_ for
DataSHIELD.

### 2.1. Connect to an Opal Server
#### Setup
For instructions on how to set up a local server for DataSHIELD, see the 
following vignette:

```{r}
vignette("deploy-local-datashield-server-with-opal")
```

Here we will use OBiBa's Opal demo server: https://opal-demo.obiba.org/ which
can be accessed with the following login credentials:

```{r}
# define global variables
## Opal server access
USERNAME <- "administrator"
USERPASS <- "password"
SERVER <- "https://opal-demo.obiba.org"
## Credentials for `dsuser`
DSUSERPASS <- "P@ssw0rd"
```

Next, define global variables used in generating the RO-Crate, such as project 
name, table references (within the project) and user identifiers.

```{r}
## Five safes variables
PEOPLE <- "dsuser"
PROJECT <- "CNSIM"
TABLES <- c("CNSIM1")
```

#### Open connection

Once the credentials and Five Safes variables are configured,we can start a 
new session to the opal server with the following command:

```{r}
# login to local server with `USERNAME` and `USERPASS`.
o <- opalr::opal.login(
  username = USERNAME,
  password = USERPASS,
  url = SERVER
)

print(o)
```

### 2.2. Create a basic RO-Crate

To create a basic RO-Crate, we will use the 
[`{rocrateR}`](https://github.com/ResearchObject/ro-crate-r) package. This 
package can be installed with the following command:

```{r, eval = FALSE, echo = TRUE}
# install.packages("pak")
pak::pak("ResearchObject/ro-crate-r@dev")
```

```{r, eval = FALSE, echo = FALSE}
# NOTE: this will only be able when the package is on the CRAN!!
# install.packages("pak")
pak::pak("rocrateR")
```

Then, a basic RO-Crate can be created with the following command:

```{r}
basic_rocrate <- rocrateR::rocrate_5s()
```

Note that this RO-Crate uses the [5s-crate](https://trefx.uk/5s-crate/0.4)
profile.

```{r}
print(basic_rocrate)
```

### 2.3. Add _Five Safes_ Elements
#### Safe Data
To add details for Safe Data, use the function `dsROCrate::safe_data()`.

```{r, out.lines=-(rocrate_lines(basic_rocrate) - 2)}
basic_rocrate <- o |>
  dsROCrate::safe_data(rocrate = basic_rocrate,
                       project = PROJECT,
                       tables = TABLES)

print(basic_rocrate) # note that the output will be truncated
```

#### Safe Project
To add details for Safe Project, use the function `dsROCrate::safe_project()`.

```{r, out.lines=-(rocrate_lines(basic_rocrate) - 2)}
basic_rocrate <- o |>
  dsROCrate::safe_project(rocrate = basic_rocrate,
                          project = PROJECT)

print(basic_rocrate) # note that the output will be truncated
```

#### Safe People
To add details for Safe People, use the function `dsROCrate::safe_people()`.

```{r safe_people, out.lines=-(rocrate_lines(basic_rocrate) + 1)}
basic_rocrate <- o |>
  dsROCrate::safe_people(rocrate = basic_rocrate, user = PEOPLE)

print(basic_rocrate) # note that the output will be truncated
```

#### Safe Setting
To add details for Safe Setting, use the function `dsROCrate::safe_setting()`.

**⚠️NOTE:** The `dsROCrate::safe_setting` function requires administrator
privileges, so here, we will have to log in with administrator credentials (if
you used a non-administrator account previously).

```{r, eval = FALSE}
# close previous connection
opalr::opal.logout(o)

# open new connection as administrator
o <- opalr::opal.login(
  username = "administrator",
  password = "password",
  url = SERVER
)
```

Then, we can proceed as per usual:
```{r, out.lines=-(rocrate_lines(basic_rocrate) - 2)}
basic_rocrate <- o |>
  dsROCrate::safe_setting(rocrate = basic_rocrate)

print(basic_rocrate) # note that the output will be truncated
```

#### Safe Outputs
To add details for Safe Outputs, use the function `dsROCrate::safe_output()`.
Currently, only log files from the operations executed by the user within a 
specific period. Set the period using `logs_from` and `logs_to`. 

**⚠️NOTE:** Similar to `dsROCrate::safe_setting`, the `dsROCrate::safe_output` 
function requires of administrator rights, so here, we will have to log in with 
administrator credentials:

```{r, eval = FALSE}
# close previous connection
opalr::opal.logout(o)

# open new connection as administrator
o <- opalr::opal.login(
  username = "administrator",
  password = "password",
  url = SERVER
)
```

---------

##### DataSHIELD operations

**⚠️NOTE:** Before extracting logs, ensure there is recent activity on the 
server for testing purposes. This can be done using the following commands:

###### Setup
You will need the following packages:

```{r, eval = FALSE}
pak::pak("DSI")
pak::pak("DSOpal")
pak::pak("dsBaseClient")
```

###### Open connection
```{r}
# run some test commands with dsBaseClient
## needed to defined the OpalDriver class in the current environment
DSOpal::Opal()
## create new login object, note that here we use the `dsuser`
builder <- DSI::newDSLoginBuilder()
builder$append(server = "study1",
               url = SERVER,
               user = "dsuser",
               password = DSUSERPASS,
               driver = "OpalDriver")
logindata <- builder$build()
conns <- DSI::datashield.login(logins = logindata)
```

###### Simulate some operations
```{r}
## assign data
DSI::datashield.assign.table(conns["study1"], 
                             symbol = "dsROCrate_test",
                             table = paste0(PROJECT, ".", TABLES[1]),
                             errors.print = TRUE)

dsBaseClient::ds.ls(datasources = conns["study1"])
dsBaseClient::ds.summary("dsROCrate_test")
```

---------

Then, we can proceed as per usual:

```{r}
basic_rocrate <- o |>
  dsROCrate::safe_output(rocrate = basic_rocrate,
                         logs_to = Sys.time())
```

```{r, out.lines=-(rocrate_lines(basic_rocrate) + 1)}
print(basic_rocrate) # note that the output will be truncated
```

#### 2.(n-1). Close connection

```{r}
opalr::opal.logout(o)
```

#### 2.n. Bag/Save RO-Crate

The resulting RO-Crate can be stored into an RO-Crate bag/archive with the
function `rocrateR::bag_rocrate`:

```{r}
# create temp directory
dir.create("./rocrates", showWarnings = FALSE)
```

NOTE: In the above example, a `path` to store the logs wasn't provided when 
calling `dsROCrate::safe_output`, before creating an RO-Crate bag, we should 
save the contents of this file first

```{r}
logs_entity <- basic_rocrate |>
  rocrateR::get_entity(type = "File")
# write file using the path given by `@id`
writeLines(
  logs_entity[[1]]$content[[1]], 
  file.path("./rocrates", logs_entity[[1]]$`@id`)
)
# remove the section `content`
logs_entity[[1]]$content <- NULL
# update the RO-Crate
basic_rocrate <- basic_rocrate |>
  rocrateR::add_entity(logs_entity[[1]], overwrite = TRUE)
```

```{r}
# create RO-Crate bag
path_to_rocrate_bag <- basic_rocrate |>
  rocrateR::bag_rocrate(path = "./rocrates")
```

We can explore the contents with the following commands:

```{r}
# extract files in temporary directory
path_to_rocrate_bag |>
  # extract contents in the same directory where the bag was stored
  rocrateR::unbag_rocrate(quiet = TRUE) |>
  # create tree with the files
  fs::dir_tree()
```

#### 2.(n+1). Clean working directory

```{r}
unlink("./rocrates", recursive = TRUE, force = TRUE)
```
